---
source: Rmd
title: "Linear regression with one continuous and one categorical explanatory variable"
objectives:
  - Explore the relationship between a continuous dependent variable and two explanatory variables, one continuous and one categorical, using ggplot2. 
  - Fit a linear regression model with one continuous explanatory variable and one categorical explanatory variable using lm(). 
  - Use the jtools package to interpret the model output. 
  - Use the interactions package to visualise the model output.
keypoints:
  - A scatterplot, with points coloured by the levels of a categorical variable, can be used to explore the relationship between two continuous variables and a categorical variable.
  - The categorical variable can be added to the `formula` in `lm()` using a `+`.
  - The model output shows separate intercepts for the levels of the categorical variable. The slope across the levels of the categorical variable is held constant. 
  - Parallel lines can be added to the exploratory scatterplot to visualise the linear regression model. 
questions:
  - How can we visualise the relationship between three variables, two of which are continuous and one of which is categorical, in R?
  - How can we fit a linear regression model to this type of data in R?
  - How can we obtain and interpret the model parameters?
  - How can we visualise the linear regression model in R?
teaching: 15
exercises: 20
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("01-")

library(dplyr)
library(tidyr)
library(ggplot2)
library(jtools)
library(interactions)
```

In this episode we will learn to fit a linear regression model when we have two explanatory variables: one continuous and one categorical. Before we fit the model, we can explore the relationship between our variables graphically. We are asking the question: does the relationship between the continuous explanatory variable and the response variable differ between the levels of the categorical variable?


## Exploratory plots
Let us take the response variable `BMI`, the continuous explanatory variable `Weight` and the categorical explanatory variable `Sex` as an example. The code below subsets our data for individuals who are older than 17 years with `filter()`. The plotting is then initiated using `ggplot()`. Inside `aes()`, we select the response variable with `y = BMI`, the continuous explanatory variable with `x = Weight` and the categorical explanatory variable with `colour = Sex`. As a consequence of the final argument, the points produced by `geom_point()` are coloured by `Sex`. Finally, we include opacity using `alpha = 0.2`, which allows us to distinguish areas dense in points from sparser areas. Note that RStudio returns the error "Removed 320 rows containing missing values (geom_point)." - this reflects that 320 participants had missing BMI and/or Weight data. 

The plot suggests that on average, participants classed as female have a higher `BMI` than participants classed as male, for any particular value of `Weight`. We might therefore want to account for this in our model by having *separate intercepts* for the levels of `Sex`.

```{r BMI_Weight_Sex exploratory plot, warning = FALSE}
dat %>%
  filter(Age > 17) %>%
  ggplot(aes(x = Weight, y = BMI, colour = Sex)) +
  geom_point(alpha = 0.2) 
```

> ## Exercise  
> You have been asked to model the relationship between `FEV1`
> and `Age` in the NHANES data, splitting observations by whether
> participants are active or ex smokers (`SmokeNow`). Use the `ggplot2`
> package to create an exploratory plot, ensuring that:
> 1. NAs are discarded from the `SmokeNow` variable.  
> 2. FEV1 (`FEV1`) is on the y-axis and Age (`Age`) is on the x-axis.
> 3. These data are shown as a scatterplot, with points coloured by 
> `SmokeNow` and an opacity of 0.4.
>
> > ## Solution
> > 
> > ```{r FEV1 vs Age by SmokeNow plot, warning = FALSE}
> > dat %>%
> >  drop_na(SmokeNow) %>%
> >  ggplot(aes(x = Age, y = FEV1, colour = SmokeNow)) +
> >  geom_point(alpha = 0.4)
> > ```
> {: .solution}
{: .challenge}

## Fitting and interpreting a multiple linear regression model
We fit the model using `lm()`. With `formula = BMI ~ Weight + Sex` we specify that our model has two explanatory variables, `Weight` and `Sex`. 

```{r fit BMI_Weight_Sex}
BMI_Weight_Sex <- dat %>%
  filter(Age > 17) %>%
  lm(formula = BMI ~ Weight + Sex)
```

The linear model equation associated with this model has the general form

$$E(y) = \beta_0 + \beta_1 \times x_1 + \beta_2 \times x_2$$.  

Notice that we now have the additional parameters, $\beta_2$ and $x_2$, in comparison to the simple linear regression model equation. In the context of our model, $\beta_1$ is the parameter for the effect of `Weight` and $\beta_2$ is the parameter for the effect of `Sex`. Recall from the simple linear regression lesson that a categorical variable has a baseline level in R. The parameter associated with the categorical variable then estimates the difference in the outcome variable in a group different from the baseline. Since "f" precedes "m" in the alphabet, R takes `female` as the baseline level. Therefore, $\beta_2$ estimates the difference in BMI between males and females, given a particular value for `Weight`. $\beta_2$ is only included when $x_2 = 1$, i.e. when the `Sex` of a participant is `male`. This results in separate intercepts for the levels of `Sex`: the intercept for `females` is given by $\beta_0$ and the intercept for `males` is given by $\beta_0 + \beta_2$. In this model, the effect of `Weight` is given by $\beta_1$, irrespective of the `Sex` of a participant. This results in equal slopes across the levels of `Sex`. 

The model parameters can be obtained using `summ()` from the `jtools` package. We include 95% confidence intervals for the parameter estimates using `confint = TRUE` and three digits after the decimal point using `digits = 3`. The intercept when `Sex = female` equals $5.538$ and the intercept when `Sex = male` equals $5.538 - 4.202 = 1.336$. Notice that in this model, the effect of `Weight` is $0.310$, regardless of `Sex`. The model equation can therefore be updated to be:

$$E(\text{BMI}) = 5.538 + 0.310 \times \text{Weight} -4.202 \times x_2,$$  

where $x_2 = 1$ for `Sex = male` and $0$ otherwise. 


```{r summ BMI_Weight_Sex}
summ(BMI_Weight_Sex, confint = TRUE, digits = 3)
```

> ## Exercise  
> 1. Using the `lm()` command, fit a multiple linear regression of FEV1
> (`FEV1`) as a function of Age (`Age`), grouped by `SmokeNow`.
> Name this `lm` object `FEV1_Age_SmokeNow`.  
> 2. Using the `summ` function from the `jtools` package, answer the following questions:
>   
> A) What `FEV1` does the model predict, on average,
> for an individual, belonging to the baseline level of `SmokeNow`,
> with an `Age` of 0?  
> B) What is R taking as the baseline level of `SmokeNow` and why?  
> C) By how much is `FEV1` expected to differ, on average, between the baseline and
> the alternative level of `SmokeNow`?  
> D) By how much is `FEV1` expected to differ, on average, for
> a one-unit difference in `Age`?  
> E) Given these values and the names of the response and explanatory
> variables, how can the general equation $E(y) = \beta_0 + {\beta}_1 
> \times x_1 + {\beta}_2 \times x_2$ be adapted to represent the model? 
>
> > ## Solution
> > 
> > ```{r FEV1 vs Age by SmokeNow model}
> > FEV1_Age_SmokeNow <- dat %>%
> >   lm(formula = FEV1 ~ Age + SmokeNow)
> > 
> > summ(FEV1_Age_SmokeNow, confint = TRUE, digits=3)
> > ```
> > 
> > A) 4928.155 mL  
> > B) `No`, because "n" precedes "y" in the alphabet.  
> > C) On average, individuals from the two categories are expected to differ
> > by 249.305 mL.  
> > D) It is expected to decrease by 35.686 mL.  
> > E) $E(\text{FEV1}) = 4928.155 - 35.686 \times \text{Age} - 249.305 \times x_2$,  
> > where $x_2 = 1$ for current smokers and 
> > $0$ otherwise.
> {: .solution}
{: .challenge}

## Visualising a multiple linear regression model
The model can be visualised using two lines, representing the levels of `Sex`. Rather than using `effect_plot()` from the `jtools` package, we use `interact_plot` from the `interactions` package. We specify the model by its name, the continuous explanatory variable using `pred = Weight` and the categorical explanatory variable using `modx = Sex`. We include the data points using `plot.points = TRUE` and introduce opacity into the data points using `point.alpha = 0.2`. 

Note that R returns the warning "Weight and Sex are not included in an interaction with one another in the model." - we will learn what interactions are and when they might be appropriate in the next episode. In this scenario we can safely ignore the warning.

```{r effect_plot BMI_Weight_Sex, warning = FALSE}
interact_plot(BMI_Weight_Sex, pred = Weight, modx = Sex,
              plot.points = TRUE, point.alpha = 0.2)
```


> ## Exercise  
> To help others interpret the `FEV1_Age_SmokeNow` model, produce a figure. 
> Make this figure using the `interactions` package.
>
> > ## Solution
> > ```{r plot FEV1_Age_SmokeNow, warning = FALSE}
> > interact_plot(FEV1_Age_SmokeNow, pred = Age, modx = SmokeNow,
> >               plot.points = TRUE, point.alpha = 0.4)
> > ```
> {: .solution}
{: .challenge}


