---
source: Rmd
title: "Making predictions from a multiple linear regression model"
objectives:
  - Calculate a prediction from a simple linear regression model using parameter estimates given by the model output.
  - Use the predict function to generate predictions from a multiple linear regression model.
keypoints:
  - Predictions of the mean in the outcome variable can be manually calculated using the model's equation.
  - Predictions of multiple means in the outcome variable alongside 95% CIs can be obtained using the `make_predictions()` function. 
questions:
  - How can predictions be calculated using the model equation of a multiple linear regression model?
  - How can R be used to obtain predictions from a multiple linear regression model?
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("03-")
library(dplyr)
library(jtools)
```

As with the simple linear regression model, the multiple linear regression model allows us to make predictions. First we will calculate predictions using the model equation. Then we will see how R can calculate predictions for us using the `predict()` function.

## Calculating predictions manually

Let us use the `BPSysAve_Age_Sex` model from the previous episode. The equation for this model was $E(\text{BPSysAve}) = 93.39 + 0.55 \times \text{Age} + 17.06 \times \text{SexMale} - \text{SexMale} \times 0.28 \times \text{Age}$, where $\text{SexMale} = 1$ for male participants and $0$ otherwise. 

For a 30-year old female, the model predicts an average systolic blood pressure of $E(\text{BPSysAve}) = 93.39 + 0.55 \times 30 + 17.06 \times 0 - 0 \times 0.28 \times 30 = 93.39 + 16.5 = 109.89$.  
For a 30-year old male, the model predicts an average systolic blood pressure of $E(\text{BPSysAve}) = 93.39 + 0.55 \times 30 + 17.06 \times 1 - 1 \times 0.28 \times 30 = 93.39 + 16.5 + 17.06 - 8.4 = 118.55$.

>## Exercise
>Given the `summ` output from our `Hemoglobin_Age_Sex` model, 
>the model can be described
>as $E(\text{Hemoglobin}) = 13.29 + 0.00 \times \text{Age} + 2.76 \times \text{SexMale} - \text{SexMale} \times 0.02 \times \text{Age}$, where $\text{SexMale} = 1$ for participants of the male `Sex` and $0$ otherwise. 
>What level of Hemoglobin does the model predict, on average, 
>for an individual of the female `Sex` with an `Age` of 40? 
> And what Hemoglobin is predicted for a male of the same `Age`?
> > ## Solution
> > For 40-year old females: $13.29 + 0.00 \times 40 + 2.76 \times 0 - 0 \times 0.02 \times 40 = 13.29$.  
> > For 40-year old males: $13.29 + 0.00 \times 40 + 2.76 \times 1 - 1 \times 0.02 \times 40 = 13.29 + 2.76 - 0.8 = 15.25$.
> {: .solution}
{: .challenge}

## Making predictions using `make_predictions()`

Using the `make_predictions()` function brings two advantages. First, when calculating multiple predictions, we are saved the effort of inserting multiple values into our model manually and doing the calculations. Secondly, `make_predictions()` returns 95% confidence intervals around the predictions, giving us a sense of the uncertainty around the predictions. 

To use `make_predictions()`, we need to create a `tibble` with the explanatory variable values for which we wish to have mean predictions from the model. We do this using the `tibble()` function. Note that the column names must correspond to the names of the explanatory variables in the model, i.e. `Age` and `Sex`. In the code below, we create a `tibble` with `Age` having the values 30, 40, 50 and 60 for females and males. We then provide `make_predictions()` with this `tibble`, alongside the model from which we wish to have predictions. By default, `make_predicitons()` returns
95% confidence intervals for the mean predictions. 

From the output we can see that the model predicts an average systolic blood pressure of 109.8596 for a 30-year old female. The confidence interval around this prediction is [109.0593, 110.6599]. 

```{r predict BPSysAve_Age_Sex}
BPSysAve_Age_Sex <- dat %>%
  filter(Age > 17) %>%
  lm(formula = BPSysAve ~ Age * Sex)

predictionDat <- tibble(Age = c(30, 40, 50, 60,
                                30, 40, 50, 60),
                        Sex = c("female", "female", "female", "female",
                                "male", "male", "male", "male"))

make_predictions(BPSysAve_Age_Sex, new_data = predictionDat)
```


>## Exercise
>Using the `make_predictions()` function, obtain the expected mean Hemoglobin levels
>predicted by the `Hemoglobin_Age_Sex` model for individuals with an Age 
> of 20, 30, 40 and 50. Obtain confidence intervals for these predictions. 
How are these confidence intervals interpreted?
> > ## Solution
> > ```{r predict from Hemoglobin_Age_Sex}
> > Hemoglobin_Age_Sex <- dat %>%
> >  filter(Age > 17) %>%
> >  lm(formula = Hemoglobin ~ Age * Sex)
> > 
> > predictionDat <- tibble(Age = c(20, 30, 40, 50,
> >                                  20, 30, 40, 50),
> >                          Sex = c("female", "female", "female", "female",
> >                                  "male", "male", "male", "male"))
> > 
> > make_predictions(Hemoglobin_Age_Sex, new_data = predictionDat)
> > ```
> > Recall that 95% of the 95% confidence intervals are expected to contain the 
> > population mean. 
> > Therefore, we can be fairly confident that the true population means lie 
> > somewhere between the bounds of the intervals, assuming that our model is good.
> {: .solution}
{: .challenge}


