---
source: Rmd
title: "Assessing multiple linear regression model fit and assumptions"
objectives:
  - Use the adjusted R squared value as a measure of model fit.
  - Assess whether the assumptions of the multiple linear regression model have been violated.
keypoints:
questions:
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("04-")
library(dplyr)
library(jtools)
library(interactions)
library(ggplot2)
library(patchwork)
```

In this episode we will check the fit and assumptions of multiple linear 
regression models. We will use the extension to $R^2$ for multiple linear 
regression models, the adjusted $R^2$ ($R^2_{adj}$). We will also learn
to assess the six linear regression assumptions in the case of 
multiple linear regression.

## Measuring fit using $R^2_{adj}$
In the simple linear regression [episode](https://carpentries-incubator.github.io/simple-linear-regression-public-health/06-fitAndAssumptionSLR) 
on model assumptions we learned that $R^2$ quantifies 
the proportion of variation in the outcome variable explained by the 
explanatory variable. In the context of multiple linear regression, a
caveat emerges to $R^2$: adding explanatory variables to our model will
almost always increase $R^2$, even if the explanatory variables do not
have relationships with the response variable. Therefore, when interpreting
the output from `summ()`, look at the $R^2_{adj}$. This is an $R^2$ corrected
for the number of variables in the model. In some cases, the adjusted and
unadjusted metrics will be (near) equal. In other cases, the differences
will be larger. 

>## Exercise
>Find the adjusted R-squared value for the `summ` output of our `Hemoglobin_Age_Sex` model from 
>[episode 2](carpentries-incubator.github.io/multiple-linear-regression-public-health/02-contAndCatInt/).
>What proportion of variation in hemoglobin is explained by age, sex and their interaction in our model? 
>Does our model account for most of the variation in `Hemoglobin`?
> > ## Solution
> > ```{r R-squared challenge}
> > Hemoglobin_Age_Sex <- dat %>%
> >   filter(Age > 17) %>%
> >   lm(formula = Hemoglobin ~ Age * Sex)
> > 
> > summ(Hemoglobin_Age_Sex, confint = TRUE, digits = 3)
> > ```
> > 
> > Since $R^2_{adj} = 0.339$, our model accounts for approximately 34% of the variation in
> > `Hemoglobin`. Our model explains 34% of the variation in `Hemoglobin`, which
> > a model that always predicts the mean would not. 
> {: .solution}
{: .challenge}

## Assessing the assumptions of the multiple linear regression model
We learned to assess six assumptions in the simple linear regression [episode](https://carpentries-incubator.github.io/simple-linear-regression-public-health/06-fitAndAssumptionSLR) 
on model assumptions. These assumptions also hold in the multiple linear 
regression context, although in some cases their assessment is a little more 
extensive. Below we will practice assessing these assumptions. 

### Validity
Recall that the validity assumption states that the model is appropriate 
for the research question. Validity is assessed through three questions:

A) Does the outcome variable reflect the *phenomenon of interest*?  
B) Does the model include *all relevant explanatory variables*?     
C) Does the model generalise to our *case of interest*?  

>## Exercise
> You are asked to use the `FEV1_Age_SmokeNow` model to answer the following research question:  
> 
> "Does the effect of age on FEV1 differ by smoking status in US adults?"  
> 
> Using the three points above, assess the validity of this model for
> the research question. 
> > ## Solution
> A) The research question is on FEV1, which is our outcome variable. Therefore our outcome variable reflects the phenomenon of interest.   
> B) Our research question asks whether the effect of age differs by smoking status, which can be tested using an interaction between `Age` and `SmokeNow`. Since our model does not include this interaction, our model does not include all relevant explanatory variables.    
> C) Since the NHANES data was collected from individuals in the US, our model should generalise to our case of interest.  
> {: .solution}
{: .challenge}

### Representativeness
Recall that the representativeness assumption states that the sample is 
representative of the population to which we are generalising our findings. This
assumption is assessed in the same way as for the case of the simple linear
regression model, so we will go through another exercise at this point. 

### Linearity and additivity
Recall that this assumption states that our outcome variable has a linear, 
additive relationship with the explanatory variables. 

The linearity component is assessed in the same way as in the simple
linear regression case. For example, recall that the relationship between
combined diastolic blood pressure (`BPDiaAve`) and age in months (`AgeMonths`)
was curved. Adding a squared term to the model helped us to model this
non-linear relationship. We can add a `Sex` term to fit separate
non-linear curves to this data:

```{r non-linear interaction BPDiaAve AgeMonths Sex, warning = FALSE}
BPDiaAve_AgeMonthsSQ_Sex <- lm(BPDiaAve ~ AgeMonths + I(AgeMonths^2) + Sex, data = dat)
interact_plot(BPDiaAve_AgeMonthsSQ_Sex, pred = AgeMonths, modx = Sex,
            plot.points = TRUE, interval = TRUE,
            point.size = 0.7) +
  ylab("Combined diastolic blood pressure") +
  xlab("Age in Months") 

```

The non-linear relationship has now been modeled separately for levels
of a categorical variable. 

Recall that the additivity component of the linearity and additivity assumption 
means that the effect of any explanatory variable on the outcome variable does 
not depend on another explanatory variable in the model. In other words, that 
our model includes all necessary interactions between explanatory variables. 
In the above example, the additivity component does not appear to be violated:
it does not look like the data requires an interaction between age and sex. 
However, to be sure of this we may compare the model fit of models with and
without the interaction using the tools discussed in this episode.

>## Exercise
> In the example above we saw that a model with a squared explanatory variable 
> can also include separate intercepts for levels of a categorical variable.  
>
> Recall out `child_logWeight_Height_lm` model from the previous lesson,
> which modeled the relationship between the log of child weight
> and child height:
> 
> ```{r non-linearity challenge intro, warning = FALSE, message = FALSE}
> child_logWeight_Height_lm <- dat %>%
>   filter(Age < 18) %>%
>   lm(formula = log(Weight) ~ Height)
> 
> effect_plot(child_logWeight_Height_lm, pred = Height, 
>                   plot.points = TRUE, interval = TRUE,
>                   colors = c("red")) 
> ```
>
> Extend this model by adding separate intercepts 
> for the levels of the `Sex` variable. 
> 
> > ## Solution
> > ```{r non-linearity challenge part 2, warning = FALSE, message = FALSE}
> > child_logWeight_Height_Sex <- dat %>%
> >   filter(Age < 18) %>%
> >   lm(formula = log(Weight) ~ Height + Sex)
> > 
> > interact_plot(child_logWeight_Height_Sex, pred = Height, modx = Sex,
> >               plot.points = TRUE, interval = TRUE,
> >               point.size = 0.7)
> > ```
> > 
> {: .solution}
{: .challenge}

### Independent errors
Recall that when observations in our data are grouped, this can result in a violation of the independent errors assumption. If there are a few levels in our grouping variable (say, less than 6) then we may overcome the violation by including the grouping variable as an explanatory variable in our model. If the grouping variable has many more levels, we may opt for a modelling approach more complex than multiple linear regression. 

>## Exercise
> In which of the following scenarios are we at risk of violating the independent errors assumption? In those cases, should we be working with an extra explanatory variable?
> 
> A) We are modelling the effect of average daily calorie intake on BMI in the adult UK population. We have one observation per participant and participants are known to belong to one of five income brackets.  
> B) We are modelling the effect of age on grip strength in adult females in the UK. Whether participants are physically active is known.  
> C) We are asking whether people's sprinting speed is increased after partaking in an athletics course. Our data includes measurements of 100 participant's sprinting speed before and after the course. 
> 
> > ## Solution
> > A) Since we have multiple observations per income bracket, our observations are not independent. There are five levels in the income bracket variable, so we may chose to include income bracket as an explanatory variable.  
> > B) Since we have multiple observations per level of physical activity, our observations are not independent. Since there are two levels in our physical activity variable, we could use physical activity as an explanatory variable in our model.   
> > C) This data has two levels of grouping: within individuals (two measurements per individual) and timing (before/after the course). While timing can be included as an explanatory variable (two levels), it would not be appropriate to include individuals as an explanatory variable (100 levels). In this scenario we would opt for a more complex modelling approach (a mixed effect model).  
> {: .solution}
{: .challenge}

```{r}
residualData <- tibble(resid = resid(BPSysAve_Age_Sex),
                    fitted = fitted(BPSysAve_Age_Sex),
                    height = BPSysAve_Age_Sex$model$Age,
                    sex = BPSysAve_Age_Sex$model$Sex)

p1 <- ggplot(residualData, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.1) +
  geom_smooth()

p2 <- ggplot(residualData, aes(x = height, y = resid)) +
  geom_point(alpha = 0.1) +
  geom_smooth()

p3 <- ggplot(residualData, aes(x = sex, y = resid)) +
  geom_violin() + 
  geom_jitter(alpha = 0.1, width = 0.2) 


p1 + p2 + p3
```
